\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-11 paper.tex
added neuroscience citations

Viewed as a probabilistic inference problem,
our goal is to jointly infer a single DSL (written $\mathcal{D}$)
and many programs (written $p$), one for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
inference
scheme~\cite{le2016inference}.
 The neural recognition model and the
generative model embodied in the DSL jointly train each other, as they
iteratively learn to solve more programming tasks.

@@

 Our goal is to induce a DSL while finding programs solving each of the
tasks.  We take inspiration from the Wake/Sleep
algorithm~\cite{hinton1995wake}, as well as the
Exploration-Compression algorithm for bootstrap
learning~\cite{Dechter:2013:BLV:2540128.2540316}.  Wake/Sleep
alternates between training a generative model on samples from a
recognition model (our $q$), and training a recognition model on
samples from the generative model (our $\mathcal{D},\theta$).  EC
alternates between exploring the space of solutions to a set of tasks,
and compressing those solutions to suggest new search primitives
for the next exploration stage.
We combine these ideas into an inference strategy that iterates
through three steps: a \textbf{Wake} cycle uses the current DSL and
recognition model to search for programs that solve the tasks.  The
\textbf{Sleep-G} and \textbf{Sleep-R} cycles update the DSL and the
recognition model, respectively.  Crucially, these steps bootstrap 
each off other (Fig.~\ref{feeding}):

\\\noindent \textbf{Wake: Searching for programs.}  Our program search
is informed by both the DSL and the recognition model. When these
improve, we find more programs solving the tasks.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce the DSL
from the programs found in the wake phase, aiming
to maximally compress (or, raise the prior probability of) these
programs.  As we solve more tasks, we hone in on richer DSLs that
more closely match the domain.
\\\noindent \textbf{Sleep-R: Learning a neural recognition model.}  We
update the recognition model training on two data sources: samples
from the DSL (as in the Helmholtz Machine's ``sleep'' phase), and
programs found by the search procedure during waking. As the DSL
improves and as search finds more programs, the recognition model gets
both more data to train on, and better data.

@@

Sec.~\ref{mathematicalFraming} frames this 3-step procedure as
a means of maximizing a lower bound on the posterior probability of the DSL given the tasks.
Sec.~\ref{explorationSection} explains how we search for programs that solve the tasks;
Sec.~\ref{recognitionSection} explains how we train a neural network to accelerate program search; and
Sec.~\ref{grammarInductionSection} explains how we induce a DSL from programs.

@@

For each domain,
we are interested both in how many tasks the
agent can solve and how quickly it can find those solutions.
For symbolic regression, we also care about the quality
of the solution, as measured by the likelihood model $\probability[x|p]$,
e.g. did the agent correctly explain a linear function
using two numbers, or did it introduce extraneous parameters?
Tbl.~\ref{baselineComparisons}
compares our model against these baselines.
Our full model consistently
improves on the baselines,
sometimes dramatically (string editing and symbolic regression).
The recognition model 
consistently increases the number of solved held-out tasks,
and lesioning it also slows down the convergence of the algorithm,
taking more wake/sleep cycles to reach a given number of tasks solved (Fig.~\ref{learningCurves}).
This supports our view of the recognition model as a way of accelerating the search over programs.

@@

 \section{Contribution and Outlook}
We contribute a model, \systemEnding, that learns to program by bootstrapping a DSL with new
domain-specific primitives that the algorithm itself discovers,
together with a neural recognition model that learns how to
efficiently deploy the DSL on new tasks. We believe this integration
of top-down symbolic representations and bottom-up neural networks --
both of them learned -- could help make program induction systems more
generally useful for AI. Many directions remain open. Two immediate
goals 
are to integrate more sophisticated neural recognition
models~\cite{devlin2017robustfill} and program
synthesizers~\cite{solar2008program}, which may improve performance in
some domains over the generic methods used here, and to
explore DSL meta-learning: Can we find a \emph{single} universal
primitive set that could effectively bootstrap DSLs for new domains,
including the four domains considered but many others?
\bibliography{main}
\bibliographystyle{icml2018}

>>>>

Viewed as a probabilistic inference problem,
our goal is to jointly infer a single DSL (written $\mathcal{D}$)
and many programs (written $p$), one for each set of inputs/outputs (\emph{tasks}, written $x$).
A DSL $\mathcal{D}$ is a collection of domain-specific subroutines.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a distribution over programs.
Fig.~\ref{graphicalModel}(a) diagrams this inference problem as a hierarchical Bayesian model.
Inference in this model is
difficult because the programs are unobserved,
and so we must solve a hard search problem to recover them. To make
search tractable we learn a bottom-up \emph{recognition
  model} (written $q(\cdot )$, Fig.~\ref{graphicalModel}(b)).
The recognition model $q(\cdot )$ is a neural
network that regresses from input/output pairs to a distribution over
programs likely to explain the input/outputs. We can also view $q(\cdot )$ as implementing an amortized
inference
scheme~\cite{le2016inference}.
 The neural recognition model and the
generative model embodied in the DSL jointly train each other, as they
iteratively learn to solve more programming tasks.

@@

 Our goal is to induce a DSL while finding programs solving each of the
tasks.  We take inspiration from the Wake/Sleep
algorithm~\cite{hinton1995wake}, as well as the
Exploration-Compression algorithm (EC) for bootstrap
learning~\cite{Dechter:2013:BLV:2540128.2540316}.  Wake/Sleep
alternates between training a generative model on samples from a
recognition model (our $q$), and training a recognition model on
samples from the generative model (our $\mathcal{D},\theta$).  EC
alternates between exploring the space of solutions to a set of tasks,
and compressing those solutions to suggest new search primitives
for the next exploration stage.
We combine these ideas into an inference strategy that iterates
through three steps: a \textbf{Wake} cycle uses the current DSL and
recognition model to search for programs that solve the tasks.  The
\textbf{Sleep-G} and \textbf{Sleep-R} cycles update the DSL and the
recognition model, respectively.  Crucially, these steps bootstrap 
off each other (Fig.~\ref{feeding}):

\\\noindent \textbf{Wake: Searching for programs.}  Our program search
is informed by both the DSL and the recognition model. When these
improve, we find more programs solving the tasks.
\\\noindent\textbf{Sleep-G: Improving the DSL.} We induce the DSL
from the programs found in the wake phase, aiming
to maximally compress (or, raise the prior probability of) these
programs.  As we solve more tasks, we hone in on richer DSLs that
more closely match the domain.\footnote{This is loosely biologically
    inspired by the formation of abstractions during sleep memory
    consolidation~\cite{DUDAI201520}}
\\\noindent \textbf{Sleep-R: Learning a neural recognition model.}  We
update the recognition model by training on two data sources: samples
from the DSL (as in the Helmholtz Machine's ``sleep'' phase), and
programs found by the search procedure during waking. As the DSL
improves and as search finds more programs, the recognition model gets
both more data to train on, and better data.\footnote{These two sources are also
    loosely biologically inspired by the distinct episodic replay and
    hallucination components of dream sleep~\cite{fosse2003dreaming}}

@@

Sec.~\ref{mathematicalFraming} frames this 3-step procedure as
a means of maximizing a lower bound on the posterior probability of the DSL given the tasks.
Sec.~\ref{explorationSection} explains how we search for programs that solve the tasks;
Sec.~\ref{recognitionSection} explains how we train a neural network to  search for programs; and
Sec.~\ref{grammarInductionSection} explains how we induce a DSL from programs.

@@

For each domain,
we are interested both in how many tasks the
agent can solve and how quickly it can find those solutions.
For symbolic regression, we also care about the quality
of the solution, as measured by the likelihood model $\probability[x|p]$,
e.g. did the agent correctly explain a linear function
using two numbers, or did it introduce extraneous parameters?
Tbl.~\ref{baselineComparisons}
compares our model against these baselines.
Our full model consistently
improves on the baselines,
sometimes dramatically (string editing and symbolic regression).
The recognition model 
consistently increases the number of solved held-out tasks,
and lesioning it also slows down the convergence of the algorithm,
taking more wake/sleep cycles to reach a given number of tasks solved (Fig.~\ref{learningCurves}).
This supports the view of the recognition model as a way of accelerating the search over programs.

@@

 \section{Contribution and Outlook}
We contribute a model, \systemEnding, that learns to program by bootstrapping a DSL with new
domain-specific primitives that the algorithm itself discovers,
together with a neural recognition model that learns how to
efficiently deploy the DSL on new tasks. We believe this integration
of top-down symbolic representations and bottom-up neural networks --
both of them learned -- could help make program induction systems more
generally useful for AI. Many directions remain open. Two immediate
goals 
are to integrate more sophisticated neural recognition
models~\cite{devlin2017robustfill} and program
synthesizers~\cite{solar2008program}, which may improve performance in
some domains over the generic methods used here.
Another direction is to
explore DSL meta-learning: Can we find a \emph{single} universal
primitive set that could effectively bootstrap DSLs for new domains,
including the four domains considered,  but also many others?
\bibliography{main}
\bibliographystyle{icml2018}
